Comparison of particle trajectories and collision operators for collisional transport in nonaxisymmetric plasmas

### 2014-02-27 doc/sfincsPaper/sfincsPaper.tex
Implemented Hakan's suggestions for conclusion of manuscript

Different stellarator transport codes make different assumptions about the
details of the drift-kinetic equation, and our goal in this work has been
to examine the consequences of these details.
In this work, we have compared three versions of the drift-kinetic equation
for a stellarator, consisting of (\ref{eq:kinetic4}) with
the coefficients (\ref{eq:fullTrajectories}), (\ref{eq:DKESTrajectories}),
or (\ref{eq:partialTrajectories}).
These three sets of expressions for $\vect{\dot{r}}_a$,
$\dot{x}_a$, and $\dot{\xi}_a$ may be interpreted as effective
particle trajectories (although we solve each form of the kinetic equation
using continuum numerical methods).
At first glance,  (\ref{eq:fullTrajectories}) and  (\ref{eq:partialTrajectories})
appear more accurate than  (\ref{eq:DKESTrajectories}).
However, as we have shown in section \ref{sec:sources},
the kinetic equation (\ref{eq:kinetic4})
with  (\ref{eq:fullTrajectories}) or  (\ref{eq:partialTrajectories})
is not well behaved when $E_r \ne 0$,
with an unphysical constraint forced upon the distribution function.
This analytic property of the kinetic equation must be dealt with
before attempting to solve the equation numerically.

To eliminate this problem of unphysical constraints,
we propose formulating the kinetic problem as in  (\ref{eq:blocks}) with (\ref{eq:sources}).
A particle and heat source are introduced,
along with the additional constraints that all the flux-surface-averaged
density and pressure reside in the leading-order Maxwellian.
For the model (\ref{eq:DKESTrajectories}), the sources always vanish.
For the model  (\ref{eq:fullTrajectories}), the particle source
vanishes for any $E_r$, and the energy source vanishes when $E_r$
takes on its ambipolar value.
The equations (\ref{eq:blocks}) have been implemented in a new time-independent continuum code SFINCS,
and the resulting ion transport matrices have been compared for the geometries of the LHD and W7-X stellarators.
When $E_r$ is below roughly one-third of the resonant value,
the three models give nearly indistinguishable results. 
This finding confirms that the incompressible-$\vect{E}\times\vect{B}$ trajectory
model used in some codes \cite{DKES1,DKES2} is quite accurate in this
small-$E_r$ regime, which is typically satisfied in experiments.
Physically, the effect of $E_r$ in this regime is to generate a $\sqrt{\nu}$ regime
of transport due to poloidal precession of helically trapped particles,
and this process is retained (at least approximately) in all three trajectory models.
Once $E_r$ approaches the resonance, however, the three trajectory models
yield substantially different results.
This $E_r \sim \Erres$ regime can be relevant to experiments with high ratios $\Te / \Ti$
\cite{Lore, Briesemeister} and strong gradients \cite{Baldzuhn}.
In figure \ref{fig:2species}, we find that in the large-$E_r$ region anticipated for the edge of W7-X,
the bootstrap current density in the full trajectory model is modestly reduced (by $8\%$)
compared to an incompressible-$\vect{E}\times\vect{B}$ calculation,
but should larger values of $E_*$ arise, we expect the deviation could grow more significant.

>>>>

Different stellarator transport codes make different assumptions about the
details of the drift-kinetic equation, and our goal in this work has been
to examine the consequences of these details.
In this work, we have compared three versions of the drift-kinetic equation
for a stellarator, consisting of (\ref{eq:kinetic4}) with
the coefficients (\ref{eq:fullTrajectories}), (\ref{eq:DKESTrajectories}),
or (\ref{eq:partialTrajectories}).
These three sets of expressions for $\vect{\dot{r}}_a$,
$\dot{x}_a$, and $\dot{\xi}_a$ may be interpreted as effective
particle trajectories (although we solve each form of the kinetic equation
using continuum numerical methods).
Equations  (\ref{eq:fullTrajectories}) and  (\ref{eq:partialTrajectories})
appear more accurate than  (\ref{eq:DKESTrajectories}),
\changed{and as we have shown in section \ref{sec:momentum},
the full trajectory model (\ref{eq:fullTrajectories}) is the only
one of the three models which gives the correct parallel viscous force 
and which preserves intrinsic ambipolarity in quasisymmetry.}
However, as we have shown in section \ref{sec:sources},
the kinetic equation (\ref{eq:kinetic4})
with  (\ref{eq:fullTrajectories}) or  (\ref{eq:partialTrajectories})
is not well behaved when $E_r \ne 0$,
with one or two unphysical constraints forced upon the distribution function.
This analytic property of the kinetic equation must be dealt with
before attempting to solve the equation numerically.

To eliminate this problem of unphysical constraints,
we propose formulating the kinetic problem as in  (\ref{eq:blocks}) with (\ref{eq:sources}).
A particle and heat source are introduced,
along with the additional constraints that all the flux-surface-averaged
density and pressure reside in the leading-order Maxwellian.
For the model (\ref{eq:DKESTrajectories}), the sources always vanish.
For the model  (\ref{eq:fullTrajectories}), the particle source
vanishes for any $E_r$, and the energy source vanishes when $E_r$
takes on its ambipolar value.
The equations (\ref{eq:blocks}) have been implemented in a new time-independent continuum code SFINCS,
and the resulting ion transport matrices have been compared for the geometries of the LHD and W7-X stellarators.
When $E_r$ is below roughly one-third of the resonant value,
the three models give nearly indistinguishable results. 
This finding confirms that the incompressible-$\vect{E}\times\vect{B}$ trajectory
model used in some codes \cite{DKES1,DKES2} is quite accurate in this
small-$E_r$ regime, which is typically satisfied in experiments.
Physically, the effect of $E_r$ in this regime is to generate a $\sqrt{\nu}$ regime
of transport due to poloidal precession of helically trapped particles,
and this process is retained (at least approximately) in all three trajectory models.
Once $E_r$ approaches the resonance, however, the three trajectory models
yield substantially different results.
This $E_r \sim \Erres$ regime can be relevant to experiments with high ratios $\Te / \Ti$
\cite{YokoyamaCERC, Lore, Briesemeister} and strong gradients \cite{Baldzuhn}.
In figure \ref{fig:2species}, we find that in the large-$E_r$ region anticipated for the edge of W7-X,
the bootstrap current density in the full trajectory model is modestly reduced (by $8\%$)
compared to an incompressible-$\vect{E}\times\vect{B}$ calculation,
but should larger values of $E_*$ arise, we expect the deviation could grow more significant.
